Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-08-15 supplement.tex
Section on statistics computed from whole genome

\subsection{Filtering and subsetting data}

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering three different recombination maps,
the OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn},
and a recombination map inferred from the Nama \citep{Van_Eeden2021-od}.
Statistics were largely unchanged by using a different map
(Figures~\ref{fig:supp-omni-hapmap-comparison},~\ref{fig:supp-omni-nama-comparison}).
The OMNI and HapMap recombination maps resulted in LD statistics that were
nearly identical, with only minor deviations between the OMNI/HapMap and the
Nama-inferred recombination maps. Both the OMNI and HapMap maps were inferred
using array data, which is sparser than whole-genome sequencing data as was
used in the Nama map inference. In subsequent analyses, we used data based on
recombination rates from the OMNI YRI map, and in the Supplemental Results we
show that the inferences are robust when using the Nama recombination map
(Section~\ref{sec:supp-alt-data}).

@@

@@

Throughout, we used inferred recombination maps based on the YRI OMNI array
data \citep{1000_Genomes_Project_Consortium2015-zq}. Because recombination
rates can vary between populations, we tested the robustness of our demographic
inference to the recombination map used to compute LD. We recomputed diversity
and LD statistics using recombination rates from the Nama recombination map
\citep{Van_Eeden2021-od}, and refit our best-fit demographic models to these
statistics.

@@

Among human populations, the Nama are relatively highly diverged from the
sampled populations used in estimating the OMNI recombination maps (such as the
YRI), and the two recombination maps were inferred using different
computational methods from different types of sequencing data (arrays vs. whole
genomes). Therefore, differences between these two maps are likely near the
upper bound for differences in recombination rates between present-day human
populations \citep{Van_Eeden2021-od}. Because our fits were highly consistent
when using the two recombination maps, we conclude that inferences using the
diversity and LD statistics here are robust to variation in recombination rates
between present-day populations as well as to any differences in inferred
recombination rates due to true recombination rate differences or typical
statistical error in recombination rate estimation.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-nama-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using OMNI YRI and
        Nama-specific recombination map.}
        The OMNI YRI recombination map and the Nama-inferred map
        \citep{Van_Eeden2021-od} differ in two primary ways: 1) the YRI and
        Nama have relatively highly diverged ancestries, and 2) the YRI
        map was inferred from low-density array data, while the Nama map
        was inferred from whole-genome sequencing data.
        Nonetheless, LD statistics computed under recombination rates from
        the two maps remain largely concordant, and inferences are
        unaffected by recombination map choice (Section~\ref{sec:supp-results}).
        Shading represents 95\% confidence intervals from the Nama-inferred
        map. Notations for each statistic are described in
        section~\ref{sec:supp-ld-stats}, and indexes represent populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-nama-comparison}
\end{figure}

>>>>

\subsection{Filtering and subsetting data}\label{sec:filtering}

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering three different recombination maps,
the OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn},
and a recombination map inferred from the Nama \citep{Van_Eeden2022-od}.
Statistics were largely unchanged by using a different map
(Figures~\ref{fig:supp-omni-hapmap-comparison},~\ref{fig:supp-omni-nama-comparison}).
The OMNI and HapMap recombination maps resulted in LD statistics that were
nearly identical, with only minor deviations between the OMNI/HapMap and the
Nama-inferred recombination maps. Both the OMNI and HapMap maps were inferred
using array data, which is sparser than whole-genome sequencing data as was
used in the Nama map inference. In subsequent analyses, we used data based on
recombination rates from the OMNI YRI map, and in the Supplemental Results we
show that the inferences are robust when using the Nama recombination map
(Section~\ref{sec:supp-alt-data}).

@@

Statistics were computed from intergenic loci, as described in
Section~\ref{sec:filtering}, to avoid biases due to direct or indirect
selection in protein-coding regions. To assess how strongly selection in coding
regions affects LD and diversity statistics, we also computed statistics from
all regions of the genome that pass the Thousand Genomes strict mask.
Genome-wide statistics showed slight deviations from statistics inferred from
intergenic regions alone (Figure~\ref{fig:supp-inter-all-comparison}), although
the differences were smaller than the differences between our best-fit models
and the data they were fit to. Nonetheless, we chose to use only intergenic
data to avoid possible biases due to selection in and around protein-coding
regions.

@@

Throughout, we used inferred recombination maps based on the YRI OMNI array
data \citep{1000_Genomes_Project_Consortium2015-zq}. Because recombination
rates can vary between populations, we tested the robustness of our demographic
inference to the recombination map used to compute LD. We recomputed diversity
and LD statistics using recombination rates from the Nama recombination map
\citep{Van_Eeden2022-od}, and refit our best-fit demographic models to these
statistics.

@@

Among human populations, the Nama are relatively highly diverged from the
sampled populations used in estimating the OMNI recombination maps (such as the
YRI), and the two recombination maps were inferred using different
computational methods from different types of sequencing data (arrays vs. whole
genomes). Therefore, differences between these two maps are likely near the
upper bound for differences in recombination rates between present-day human
populations \citep{Van_Eeden2022-od}. Because our fits were highly consistent
when using the two recombination maps, we conclude that inferences using the
diversity and LD statistics here are robust to variation in recombination rates
between present-day populations as well as to any differences in inferred
recombination rates due to true recombination rate differences or typical
statistical error in recombination rate estimation.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-nama-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using OMNI YRI and
        Nama-specific recombination map.}
        The OMNI YRI recombination map and the Nama-inferred map
        \citep{Van_Eeden2022-od} differ in two primary ways: 1) the YRI and
        Nama have relatively highly diverged ancestries, and 2) the YRI
        map was inferred from low-density array data, while the Nama map
        was inferred from whole-genome sequencing data.
        Nonetheless, LD statistics computed under recombination rates from
        the two maps remain largely concordant, and inferences are
        unaffected by recombination map choice (Section~\ref{sec:supp-results}).
        Shading represents 95\% confidence intervals from the Nama-inferred
        map. Notations for each statistic are described in
        section~\ref{sec:supp-ld-stats}, and indexes represent populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-nama-comparison}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-intergenic-all-snps-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed from intergenic loci and
        from all loci across the genome.} Statistics computed from intergenic
        regions (dashed lines, shown with computed standard errors as shaded
        regions) and those computed from all regions genome wide (solid lines)
        do not show large differences. We chose to perform model inference
        using intergenic regions alone to reduce bias due to selection in and
        around protein-coding regions (Section~\ref{sec:computing-stats}).
        Notations for each statistic are described in
        section~\ref{sec:supp-ld-stats}, and indexes represent populations (0:
        Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-inter-all-comparison}
\end{figure}
